Bayesian RKHS-based methods in functional regression\support{Support information of the article.}

### 2022-08-06 notes/notes.tex
Reorganize; edit paper; add mcmc notes

% Some custom commands
\newcommand{\R} {\ensuremath{\mathds{R}}}
\newcommand{\E} {\ensuremath{\mathds{E}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\bm}{\symbf}
\DeclareMathOperator{\var} {Var}
\DeclareMathOperator{\cov} {Cov}

@@

>>>>

% Some custom commands
\newcommand{\R} {\ensuremath{\mathds{R}}}
\newcommand{\E} {\ensuremath{\mathds{E}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\bm}{\symbf}

%% Scalar product
\newcommand\dotprod[2]{\left\langle #1,\, #2 \right\rangle}

\DeclareMathOperator{\var} {Var}
\DeclareMathOperator{\cov} {Cov}

@@

  \textbf{Why choose the parameter in the RKHS?}

  - Small RKHS norm implies small variations and hence a higher degree of smoothness. Specifically, \(\|f\|_K\) controls how fast the function varies over the input space with respect to the geometry defined by the kernel:
  \[
  |f(t) - f(s)| = |\dotprod{f}{K_t - K_s}_K| \leq \|f\|_K \|K_t - K_s\|_K.
  \]

  - Note that in an RKHS the evaluation functional is continuous, so a small perturbation \(f\mapsto f + \delta f\) results in a small change in \(f(x)\) for all \(x\). Not equivalence classes.

  - The model is \textit{linear} because of the correspondence between the RKHS and the linear span of the process (Loève's isometry).

### 2022-08-06 paper/paper.tex
Reorganize; edit paper; add mcmc notes

\begin{abstract}
We propose a novel Bayesian approach for functional linear and logistic regression models based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology, and the posterior distribution can be estimated from Markov Chain Monte Carlo methods. Several prediction strategies derived from this posterior distribution turn out to be competitive against other usual alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection procedure.
\end{abstract}

@@

FDA is undoubtedly an active area of research, which finds applications in a wide variety of fields, such as biomedicine, finance, meteorology or chemistry \citep[see][]{ullah2013applications}. Accordingly, there are many recent contributions on how to tackle functional data problems, both from a theoretical and practical standpoint. Chief among them is the approach of reducing the problem to a finite-dimensional one, for example using a truncated basis expansion or spline interpolation methods \citep{aguilera2013comparative}. At the same time, much effort has also been put into the opposite task: generalizing different concepts to the infinite-dimensional framework. Examples of this endeavor include a functional Mahalanobis distance \citep{berrendero2020mahalanobis} or a functional partial least squares algorithm \citep{delaigle2012methodology}. Another technique found in the related literature is the use of Gaussian processes to model the functional behavior of the data \citep[see for instance][]{shi2011gaussian}. These ideas extend the theory of Gaussian process regression in classical finite-dimensional settings \citep[e.g.][]{rasmussen2004gaussian}, providing an alternative non-parametric Bayesian approach to inference and prediction problems. Additional non-parametric methods for functional prediction and classification were notably explored in \citet{ferraty2006nonparametric}.

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will further suppose without loss of generality that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

@@

 The most common functional linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\), and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). In this case, the inference on \(\beta\) is hampered by the fact that \(L^2[0,1]\) is an extremely wide space that also contains many non-smooth or ill-behaved functions, so that any estimation procedure involving optimization on it would typically be very hard. In spite of this, model~\eqref{eq:l2-linear-model} is not flexible enough to include ``simple'' finite-dimensional models based on linear combinations of the marginals, such as \(Y=\alpha_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\); see \citet{berrendero2020general} for additional details on this. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

@@

It turns out that in both scenarios a natural alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it includes a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is especially appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018use, berrendero2018functional} for the case of functional logistic regression. Incidentally, these models also shed light on the near-perfect classification phenomenon for functional data, described by \citet{delaigle2012achieving} and further examined for example in the works of \citet{berrendero2018use} or \citet{torrecilla2020optimal}.

@@

Another set of techniques extensively studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_j)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed in the RKHS framework \citep[see for example][]{berrendero2019rkhs}, but in general they have their own dedicated algorithms and procedures. As will become apparent in the forthcoming sections, given the nature of our suggested Bayesian model we can easily isolate the marginal posterior distribution corresponding to a finite set of points \(\{t_j\}\), and thus provide a Bayesian-motivated variable selection process along with the other prediction methods that naturally arise within our model. In this way, in addition to making predictions about the input data, we can evaluate exactly which marginals of the functional explanatory variable contain the most relevant information. These points-of-impact selection models for functional predictors have also appeared recently in the literature; see \citet{poss2020superconsistent}, \citet{berrendero2016variable} or \citet{ferraty2010most} by way of illustration.

@@

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces (for a more detailed account, see for example~\citet{berlinet2004reproducing}). Let us denote by \(K(t, s)= \mathbb E[X(t)X(s)]\) the covariance function of the centered process \(X\), and in what follows suppose that it is continuous. To construct the RKHS \(\Hcal(K)\) associated with the covariance function \(K\), we start by defining the functional space \(\Hcal_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\begin{equation}\label{eq:h0}
\Hcal_0(K) = \left\{ f \in L^2[0,1]: \ f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\end{equation}
This space is endowed with the inner product \(\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j)\), given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Then, \(\Hcal(K)\) is defined to be the completion of \(\Hcal_0(K)\) under the norm induced by the scalar product \(\dotprod{\cdot}{\cdot}_K\). As it turns out, functions in this space satisfy the \textit{reproducing property} \(\dotprod{K(t, \cdot)}{f}_K = f(t)\) for all \(f \in \Hcal(K)\) and \(t \in [0, 1]\). An important consequence is that \(\Hcal(K)\) is a space of genuine functions and not of equivalence classes, since the values of the functions at particular points are in fact relevant, unlike in \(L^2\)-spaces.

@@

Lastly, a particularly useful approach in statistics is to regard \(\Hcal(K)\) as an isometric copy of a well-known space. Specifically, via \textit{Loève's isometry} one can establish a congruence \(\Psi_X\) between \(\Hcal(K)\) and the linear span of the process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) \citep[see Lemma 1.1 in][]{lukic2001stochastic}. This isometry is essentially the completion of the correspondence
  \begin{equation}\label{eq:loeves-isometry}
  \sum_{i=1}^p a_i X(t_i) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
  \end{equation}
and can be formally defined, in terms of its inverse, as \(\Psi^{-1}_X(U)(t) = \E[U X(t)]\) for \(U \in \mathcal L(X)\).
Despite the close connection between the process \(X\) and the space \(\Hcal(K)\), special care must be taken when dealing with concrete realizations of the process, since under rather general conditions the trajectories of \(X\) do not belong to the corresponding RKHS with probability one \citep[see for example][Cor.~7.1]{lukic2001stochastic}. As a consequence, the expression \(\dotprod{x}{f}_K\) is ill-defined and lacks meaning when \(x\) is a realization of \(X\). However, following Parzen's approach in his seminal work on time series analysis \citep[Th.~4E]{parzen1961approach}, we can leverage Loève's isometry and identify \(\dotprod{x}{f}_K \) with \( \Psi_x(f) := \Psi_X(f)(\omega)\), for \(x=X(\omega)\) and \(f\in \Hcal(K)\). This notation often proves to be useful and convenient.

\begin{comment}
  Explicar mejor de dónde viene la notación de Parzen, y por qué se podría pensar en usarla para expresar un producto escalar (``congruence inner product'', e.g. 4.17 en Parzen 1961b. ¿Expresión en Torrecilla et al. y Kailath?).
\end{comment}

@@

In this section we present the precise models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration are those obtained by taking a functional parameter \(\alpha \in \Hcal(K)\) and replacing the scalar product for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}. For practical and computational reasons, the value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_j \in \R, \ t_j \in [0, 1]\right\}.
\end{equation}

@@

In view of~\eqref{eq:h0p}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\beta_j\) and the times \(t_j\) separately. Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\mathbb{P}_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\Hcal(K)\) merely by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\). Even though we actually work on \(\Hcal_{0,p}(K)\), the discrete parameter \(p\) can still be selected in several meaningful ways that make use of the available data, and the set of possible values is not very large in practice.

@@

\begin{proposition}
Under the linear model~\eqref{eq:rkhs-model-linear-2}, the prior distribution implied in~\eqref{eq:prior-linear} produces the log-posterior distribution
\begin{align*}
\log \pi(\theta \mid \mathcal D_n) \propto {} & \frac{1}{2\sigma^2}\left(\|\bm Y- \alpha_0\bm{1} - \mathcal X_\tau b\|^2 + \frac{1}{g}(b - b_0)'G_\tau(b - b_0) \right)\\
& + (p+n)\log\sigma - \frac{1}{2}\log |G_\tau|,
\end{align*}
where \(\bm Y=(Y_1,\dots,Y_n)'\) and \(\bm{1}\) is an \(n\)-dimensional vector of ones.
\end{proposition}

@@

\begin{proposition}
Under the logistic model~\eqref{eq:rkhs-model-logistic-2}, the prior distribution implied in~\eqref{eq:prior-linear} produces the log-posterior distribution
\begin{align*}
  \log \pi(\theta \mid \mathcal D_n) \propto {} & \sum_{i=1}^n \left[ \left(\alpha_0 + \Psi_{X_i}(\alpha)\right)Y_i - \log\left(1 + \exp\left\{\alpha_0 + \Psi_{X_i}(\alpha)\right\}\right)\right]\\
  \quad &+ \frac{1}{2}\log |G_\tau| - p\log \sigma -\frac{1}{2g\sigma^2} (b - b_0)'G_\tau(b - b_0).
\end{align*}
Remember that \(\Psi_{X_i}(\alpha) = \sum_{j=1}^p \beta_j X_i(t_j)\).
\end{proposition}

@@

@@

\paragraph{Summarizing the posterior.} On a separate note, observe that the choice of a specific point estimator to summarize the posterior distribution results in a veiled assumption of an underlying loss function between the estimated and real parameters. In general, the mean is more sensitive to outliers and the median is more robust, but the latter assumes an \(L^1\)-type loss function while the former implicitly optimizes an \(L^2\) loss. On the other hand, the mode is also a good candidate because it represents the point of highest probability density. At any rate, these decisions are strongly dependent on several factors such as the skewness or the number of modes in the resulting posterior distribution, and thus should be made on a case-by-case basis.

>>>>

\begin{abstract}
We propose a novel Bayesian approach for functional linear and logistic regression models based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology, and the posterior distribution can be estimated through Markov chain Monte Carlo methods. Several prediction strategies derived from this posterior distribution turn out to be competitive against other usual alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection procedure.
\end{abstract}

@@

FDA is undoubtedly an active area of research, which finds applications in a wide variety of fields, such as biomedicine, finance, meteorology or chemistry \citep[see][]{ullah2013applications}. Accordingly, there are many recent contributions on how to tackle functional data problems, both from a theoretical and practical standpoint. Chief among them is the approach of reducing the problem to a finite-dimensional one, for example using a truncated basis expansion or spline interpolation methods \citep[e.g.][]{muller2005generalized, aguilera2013comparative}. At the same time, much effort has also been put into the opposite task: generalizing different concepts to the infinite-dimensional framework. Examples of this endeavor include a functional Mahalanobis distance \citep{berrendero2020mahalanobis} or a functional partial least squares algorithm \citep{delaigle2012methodology}. Another technique found in the related literature is the use of Gaussian processes to model the functional behavior of the data \citep[see for instance][]{shi2011gaussian}. These ideas extend the theory of Gaussian process regression in classical finite-dimensional settings \citep{rasmussen2004gaussian}, providing an alternative Bayesian approach to functional inference and prediction problems. Additional non-parametric methods for functional prediction and classification were notably explored in \citet{ferraty2006nonparametric}.

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will further suppose without loss of generality that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a \textit{labeled} data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

@@

 The most common functional linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\), and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). In this case, the inference on \(\beta\) is hampered by the fact that \(L^2[0,1]\) is an extremely wide space that also contains many non-smooth or ill-behaved functions, so that any estimation procedure involving optimization on it would typically be hard. In spite of this, model~\eqref{eq:l2-linear-model} is not flexible enough to include ``simple'' finite-dimensional models based on linear combinations of the marginals, such as \(Y=\alpha_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\); see \citet{berrendero2020general} for additional details on this. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory. Thus, some regularization or dimensionality reduction technique is needed for parameter estimation.

@@

It turns out that in both scenarios a natural alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown functional parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that functional space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it includes a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is especially appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018functional, berrendero2018use} for the case of functional logistic regression. Incidentally, these models also shed light on the near-perfect classification phenomenon for functional data, described by \citet{delaigle2012achieving} and further examined for example in the works of \citet{berrendero2018use} or \citet{torrecilla2020optimal}.

@@

Another set of techniques extensively studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_j)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed in the RKHS framework \citep[see for example][]{berrendero2019rkhs}, but in general they have their own dedicated algorithms and procedures. As will become apparent in the forthcoming sections, given the nature of our suggested Bayesian model we can easily isolate the marginal posterior distribution corresponding to a finite set of points \(\{t_j\}\), and thus provide a Bayesian-motivated variable selection process along with the other prediction methods that naturally arise within our model. In this way, in addition to making predictions about the input data, we can evaluate exactly which marginals of the functional explanatory variable contain the most relevant information. These points-of-impact selection models for functional predictors have also been considered in the  literature; see \citet{poss2020superconsistent}, \citet{berrendero2016variable} or \citet{ferraty2010most} by way of illustration. Another example of a related strategy is the work of \citet{james2009functional}, in which the authors propose a method to estimate \(\beta(t)\) in such a way that it is exactly zero over some regions in the domain.

@@

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces (for a more detailed account, see for example~\citet{berlinet2004reproducing}). Let us denote by \(K(t, s)= \mathbb E[X(t)X(s)]\) the covariance function of the centered process \(X\), and in what follows suppose that it is continuous. To construct the RKHS \(\Hcal(K)\) associated with the covariance function, we start by defining the functional space \(\Hcal_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\begin{equation}\label{eq:h0}
\Hcal_0(K) = \left\{ f \in L^2[0,1]: \ f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\end{equation}
This space is endowed with the inner product \(\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j)\), given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Then, \(\Hcal(K)\) is defined to be the completion of \(\Hcal_0(K)\) under the norm induced by the scalar product \(\dotprod{\cdot}{\cdot}_K\). As it turns out, functions in this space satisfy the \textit{reproducing property} \(\dotprod{K(t, \cdot)}{f}_K = f(t)\) for all \(f \in \Hcal(K)\) and \(t \in [0, 1]\). An important consequence is that \(\Hcal(K)\) is a space of genuine functions and not of equivalence classes, since the values of the functions at particular points are in fact relevant, unlike in \(L^2\)-spaces.

@@

Lastly, a particularly useful approach in statistics is to regard \(\Hcal(K)\) as an isometric copy of a well-known space. Specifically, via \textit{Loève's isometry} \citep{loeve1948fonctions} one can establish a congruence \(\Psi_X\) between \(\Hcal(K)\) and the linear span of the process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) \citep[see Lemma 1.1 in][]{lukic2001stochastic}. This isometry is essentially the completion of the correspondence
  \begin{equation}\label{eq:loeves-isometry}
  \sum_{i=1}^p a_i X(t_i) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
  \end{equation}
and can be formally defined, in terms of its inverse, as \(\Psi^{-1}_X(U)(t) = \E[U X(t)]\) for \(U \in \mathcal L(X)\).
Despite the close connection between the process \(X\) and the space \(\Hcal(K)\), special care must be taken when dealing with concrete realizations of the process, since under rather general conditions the trajectories of \(X\) do not belong to the corresponding RKHS with probability one \citep[see for example][Cor.~7.1]{lukic2001stochastic}. As a consequence, the expression \(\dotprod{x}{f}_K\) is ill-defined and lacks meaning when \(x\) is a realization of \(X\). However, following Parzen's approach in his seminal work \citep[e.g.][Th.~4E]{parzen1961approach}, we can leverage Loève's isometry and identify \(\dotprod{x}{f}_K \) with the image \( \Psi_x(f) := \Psi_X(f)(\omega)\), for \(x=X(\omega)\) and \(f\in \Hcal(K)\). This notation, viewed as a formal extension of the inner product, often proves to be useful and convenient.

@@

In this section we present the precise models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by taking a functional parameter \(\alpha \in \Hcal(K)\) and replacing the scalar product for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}. For practical and computational reasons, the value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_j \in \R, \ t_j \in [0, 1]\right\}.
\end{equation}

@@

In view of~\eqref{eq:h0p}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\beta_j\) and the times \(t_j\) separately. Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\mathbb{P}_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\Hcal(K)\) merely by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\). Even though we actually work on \(\Hcal_{0,p}(K)\), the discrete parameter \(p\) can still be selected in several meaningful ways that make use of the available data, and the set of feasible values is not very large in practice.

@@

\begin{proposition}
Under the linear model~\eqref{eq:rkhs-model-linear-2}, the prior distribution implied in~\eqref{eq:prior-linear} produces the log-posterior distribution
\begin{align*}
\log \pi(\theta \mid \mathcal D_n) \propto {} & \frac{1}{2\sigma^2}\left(\|\bm Y- \alpha_0\bm{1} - \mathcal X_\tau b\|^2 + \frac{1}{g}(b - b_0)'G_\tau(b - b_0) \right)\\
& + (p+n+2)\log\sigma - \frac{1}{2}\log |G_\tau|,
\end{align*}
where \(\bm Y=(Y_1,\dots,Y_n)'\) and \(\bm{1}\) is an \(n\)-dimensional vector of ones.
\end{proposition}

@@

\begin{proposition}
Under the logistic model~\eqref{eq:rkhs-model-logistic-2}, the prior distribution implied in~\eqref{eq:prior-linear} produces the log-posterior distribution
\begin{align*}
  \log \pi(\theta \mid \mathcal D_n) \propto {} & \sum_{i=1}^n \left[ \left(\alpha_0 + \Psi_{X_i}(\alpha)\right)Y_i - \log\left(1 + \exp\left\{\alpha_0 + \Psi_{X_i}(\alpha)\right\}\right)\right]\\
  \quad &+ \frac{1}{2}\log |G_\tau| - (p+2)\log \sigma -\frac{1}{2g\sigma^2} (b - b_0)'G_\tau(b - b_0).
\end{align*}
Remember that \(\Psi_{X_i}(\alpha) = \sum_{j=1}^p \beta_j X_i(t_j)\).
\end{proposition}

@@

\begin{comment}
LO QUE NO SIRVA, GUARDAR PARA EL TFM.

Reescribir label switching (nos pasa; lo que pasa que con p free se acentua).

Ocurre porque el likelihood es simetrico: L(theta mid y) = L(thetaperm mid y) para cualquier permutacion (hay p! posibles) --> no-identificabilidad.

"In these situations, parameter estimates and other output are inevitably aggregated
across numerous replications, and because the labels of the enumerated classes are
arbitrarily assigned, there is no guarantee that the same class would have the same
label from replication to replication."

From a Bayesian pointof view, this amounts to truncating the original prior distribution... (to verify the constraints)

As a result, in any MCMC algorithm, labels of the
components can permute multiple times between iterations of the sampler.
Paradoxically, as noted by Celeux, Hurn and Robert (2000), Fruhwirth-Schnatter (2001),
Jasra, Holmes and Stephens (2005) and Papastamoulis and Iliopoulos (2010), among others,
label switching is a prerequisite for MCMC convergence. If there is no label switching it
means that the sampler is not exploring all the modes of the posterior distribution of (1)

"When sampling from the unconstrained posterior via MCMC methods, which component of the sampled parameter corresponds to which state is unknown due to potential label switching." Si usamos la marginal para haacer inferencia --> nonsensical answers (Stephens, 2000)

Solución: añadir constraints de identificabilidad artificiales (ej relabeling, mantener ordenados, posprocesado).

Imposing an identifiability constraint on the parameters breaks the sym-
metry of the posterior distribution of the parameters, and so we might hope that it would allow us to perform sensible inference for the individual components of the mixture. Inference conditional on such a constraint may be performed by post-processing the sample (proposition 3.1 in Stephens 1997 phD thesis).

OTRA OPCION: Elegir en cada iteración como parámetro para ordenar aquel que deja mayor separación (tras escalarlos todos a [0,1]): \verb@https://helda.helsinki.fi/bitstream/handle/10138/329485/Approximate_Bayesian_computation_for_finite_mixture_models.pdf?sequence=1@

OTRA SOLUCION: "Adaptive Metropolis with Online Relabeling" --> The key step is the selection, after each proposal of a permutation (relabeling) of the proposed vector by minimizing a quadratic cost function
$L_{t-1} (x)$ over all permutations of the proposed vector, where L is defined by
$L_t(x) = (x - \mu)'\Sigma^{-1}(x - \mu)$
This step forces the posterior sample to look as unimodal as possible.

Citas: Stephens (2000), Papastamoulis \& Iliopoulos, 2010; Rodriguez \& Walker, 2014
\end{comment}

@@

\paragraph{Summarizing the posterior.} On a separate note, observe that the choice of a specific point estimator to summarize the posterior distribution results in a veiled assumption of an underlying loss function between the estimated and real parameters. In general, the mean is more sensitive to outliers and the median is more robust, but the latter assumes an \(L^1\)-type loss function while the former implicitly optimizes an \(L^2\) loss. On the other hand, the mode is also a good candidate because it represents the point of highest probability density. At any rate, these decisions are strongly dependent on several factors such as the skewness or the number of modes in the resulting posterior distribution, and thus should be made on a case-by-case basis. \incomment{Argue that, because of the label switching and multimodality, summarizing the posterior might be impractical, unhelpful, misleading, etc. So the best strategy is the posterior mean.}
